---
title: "Tick16S_Analysis_Pipeline_AllSamples_20230810"
author: "Sofia Roitman"
date: "2023-08-10"
output: html_document
---

# Set working directory
```{r}
setwd("path/to/working/directory")
```


# Load dependencies
```{r message=FALSE, error=FALSE}
# libraries and functions
library(tidyverse)
library(RVAideMemoire)
library(vegan)
library(BiocManager)
library(phyloseq) 
library(rhdf5)
library(biomformat)
library(microbiome)
library(dplyr) 
library(gcookbook)
library(tidyverse)
library(ape)
library(Biostrings) 
library(Hmisc) 
library(yaml) 
library(tidyr) 
library(stats) 
library(utils) 
library(qiime2R)
library(genefilter)
library(magrittr)
library(DESeq2)
library(plyr)
library(data.table)
library(phytools)
library(stringi)
library(ggpubr)
library(vegan)
library(microbiomeSeq)
library(microbiomeutilities)
library(ampvis2)
library(pairwiseAdonis)
library(ecole)
library(paletteer)
library(mirlyn)
library(ggtree)
library(ggplot2)
library(scales)
library(microbiomeViz)
library(miaViz)
library(scater)
library(mia)
library(mina)
library(NetCoMi)
library(devtools)
library(microeco)
library(file2meco)
library(ggh4x)

```

# Importing QIIME2 results using qiime2R
## Following: https://forum.qiime2.org/t/tutorial-integrating-qiime2-and-r-for-data-visualization-and-analysis-using-qiime2r/4121
### Creating a Phyloseq object
```{r}
# Import deblur feature table, rooted tree, and metadata file into a phyloseq object
physeq_deblur<-qza_to_phyloseq(
    features="input/_deblur-table.qza",
    tree="input/_deblur-rooted-tree.qza",
    metadata ="input/metadata.txt"
    )
physeq_deblur
## Summarize
summarize_phyloseq(physeq_deblur)

# Import taxonomy table into the phyloseq object
tax_deblur_kraken2 <- as.matrix(read.table("input/_kraken2_taxonomy.txt", sep="\t", header=T, row.names=1))
obs_new_deblur_kraken2 <- tax_table(tax_deblur_kraken2)
otus_merged_deblur_kraken2 <- merge_phyloseq(physeq_deblur,obs_new_deblur_kraken2)
## Summarize
summarize_phyloseq(otus_merged_deblur_kraken2)

```

# Clean data
```{r}
# Remove features that have only been classigied at the kingdom level
physeq_deblurK_clean <- subset_taxa(otus_merged_deblur_kraken2, Phylum != "NA")
summarize_phyloseq(physeq_deblurK_clean)

# Remove eukrayotes and archaea
physeq_deblurK_noeukarch = subset_taxa(physeq_deblurK_clean,
                                              Kingdom   != "Eukaryota",
                                              Kingdom   != "Archaea" )

summarize_phyloseq(physeq_deblurK_noeukarch)

# There are samples for which there are no observed OTUs. Let's remove those.
physeq_deblurK_nosamp50 = prune_samples(sample_sums(physeq_deblurK_noeukarch)>=50, physeq_deblurK_noeukarch)

# There are ASVs that appear only once in the data. Let's remove those as well.
physeq_deblurK_no0 = prune_taxa(taxa_sums(physeq_deblurK_nosamp50)>1, physeq_deblurK_nosamp50)



```

# Replace NAs in tax table
```{r}
taxdeK <- data.frame(tax_table(physeq_deblurK_no0))
taxdeK.clean <- data.frame(row.names = row.names(taxdeK), 
                        Kingdom = taxdeK[,1],
                        Phylum = taxdeK[,2],
                        Class = taxdeK[,3],
                        Order = taxdeK[,4],
                        Family = taxdeK[,5],
                        Genus = taxdeK[,6],
                        Species = taxdeK[,7],
                        stringsAsFactors = FALSE)
taxdeK.clean[is.na(taxdeK.clean)] <- ""

for (i in 1:7){ taxdeK.clean[,i] <- as.character(taxdeK.clean[,i])}
####### Fill holes in the tax table
taxdeK.clean[is.na(taxdeK.clean)] <- ""
for (i in 1:nrow(taxdeK.clean)){
  if (taxdeK.clean[i,2] == ""){
    kingdom <- paste("Kingdom_", taxdeK.clean[i,1], sep = "")
    taxdeK.clean[i, 2:7] <- kingdom
    } else if (taxdeK.clean[i,3] == ""){
      phylum <- paste("Phylum_", taxdeK.clean[i,2], sep = "")
      taxdeK.clean[i, 3:7] <- phylum
      } else if (taxdeK.clean[i,4] == ""){
        class <- paste("Class_", taxdeK.clean[i,3], sep = "")
        taxdeK.clean[i, 4:7] <- class
        } else if (taxdeK.clean[i,5] == ""){
          order <- paste("Order_", taxdeK.clean[i,4], sep = "")
          taxdeK.clean[i, 5:7] <- order
          } else if (taxdeK.clean[i,6] == ""){
            family <- paste("Family_", taxdeK.clean[i,5], sep = "")
            taxdeK.clean[i, 6:7] <- family
            } else if (taxdeK.clean[i,7] == ""){
              taxdeK.clean$Species[i] <- paste("Genus",taxdeK.clean$Genus[i], sep = "_")
            }
}

# Replace old taxonomy table with edited version
physeq_debKno0_taxfix <- physeq_deblurK_no0
tax_table(physeq_debKno0_taxfix) <- as.matrix(taxdeK.clean)
summarize_phyloseq(physeq_debKno0_taxfix)

saveRDS(physeq_debkno0_taxfix, "physeq_debkno0_taxfix.rds")

```

# SUBSETTING
## Create object with no Rickettsia
```{r}
no_rick = subset_taxa(physeq_debKno0_taxfix, Genus!="Rickettsia")
no_rick_prune = prune_taxa(taxa_sums(no_rick)>1, no_rick)

```

# DISTANCE MATRICES
## Bray Curtis, Unweighted UniFrac, Weighted UniFrac
```{r}
# All
bray_all <- phyloseq::distance(physeq_debKno0_taxfix, "bray")
uuni_all <- phyloseq::distance(physeq_debKno0_taxfix, "uUniFrac")
wuni_all <- phyloseq::distance(physeq_debKno0_taxfix, "wUniFrac")
```


# SIGNIFICANCE TESTING

## ALL
### PERMANOVA
```{r}
# ALL
## Bray
### Make a data frame from the otu_data
df <- data.frame(sample_data(physeq_debKno0_taxfix))
distance_methods <-c("bray_all")
set.seed(129)
# Run for loop in distance matrices         
for (i in distance_methods){ 
  form <- as.formula(paste(i, "site_sex+bb_lyme_noPN+bm_babesiosis_noPN+aph_anaplasmosis_noPN", sep="~"))
  print(form)
 adonis2(form, data=df)->result 
 print(result)
 capture.output(result, file = paste0("output/","ADONIS_bray_all",date(),".txt"))
} 


## UUNI
### Make a data frame from the otu_data
df <- data.frame(sample_data(physeq_debKno0_taxfix))
distance_methods <-c("uuni_all")
set.seed(129)
# Run for loop in distance matrices         
for (i in distance_methods){ 
  form <- as.formula(paste(i, "site_sex+bb_lyme_noPN+bm_babesiosis_noPN+aph_anaplasmosis_noPN", sep="~"))
  print(form)
 adonis2(form, data=df)->result
 print(result)
 capture.output(result, file = paste0("output/","ADONIS_uuni_all",date(),".txt"))
} 


## WUNI
### Make a data frame from the otu_data
df <- data.frame(sample_data(physeq_debKno0_taxfix))
distance_methods <-c("wuni_all")
set.seed(129)
# Run for loop in distance matrices         
for (i in distance_methods){ 
  form <- as.formula(paste(i, "site_sex+bb_lyme_noPN+bm_babesiosis_noPN+aph_anaplasmosis_noPN", sep="~"))
  print(form)
 adonis2(form, data=df)->result
 print(result)
 capture.output(result, file = paste0("output/","ADONIS_wuni_all",date(),".txt"))
} 
```

#### Posthoc
```{r}
df <- data.frame(sample_data(physeq_debKno0_taxfix))

# BRAY                 
## Pariwise Permutation MANOVAs: lyme
set.seed(129)
permanova_pairwise(bray_all,df$site_sex,permutations = 1000)

# UUNI
## Pariwise Permutation MANOVAs: sex_site
set.seed(129)
permanova_pairwise(uuni_all,df$site_sex,permutations = 1000)

# WUNI
## Pariwise Permutation MANOVAs: sex_site
set.seed(129)
permanova_pairwise(uuni_all,df$site_sex,permutations = 1000)

```

### PERMDISP
#### BRAY
```{r}
# Site sex
## make a data frame from the sample_data
sampledf <- data.frame(sample_data(physeq_debKno0_taxfix))
## Make a data frame from the otu_data
beta <- betadisper(bray_all, sampledf$site_sex)
result <- permutest(beta, pairwise = TRUE)
print(result)
capture.output(result, file = paste0("output/",i,"BETADISP_bray_all_site_sex",date(),".txt"))
### 0.001

```

#### UUNI
```{r}
# Site sex
## make a data frame from the sample_data
sampledf <- data.frame(sample_data(physeq_debKno0_taxfix))
## Make a data frame from the otu_data
beta <- betadisper(uuni_all, sampledf$site_sex)
result <- permutest(beta, pairwise = TRUE)
print(result)
capture.output(result, file = paste0("output/",i,"BETADISP_uuni_all_sitesex",date(),".txt"))
## 0.001

```

# Set theme: Use ggplot theme to make everything more legible
```{r}
ggtheme <- theme(axis.title = element_text(colour="black",family = "Helvetica",
                                           size = rel(1.5)), 
                 axis.text = element_text(family = "Helvetica",colour = "black",
                                          size = rel(1)), 
                 axis.line = element_line(size = 0.5,colour = "black"), 
                 axis.ticks = element_blank(),
                 panel.grid.major = element_line(colour="grey",size = rel(0.25)), 
                 panel.grid.minor = element_blank(), 
                 panel.background = element_blank(),
                 plot.title = element_text(colour = "black", face = "bold",
                                           size = rel(2),family = "Helvetica",hjust = 0.5,
                 geom_point(size = 20)))

sex_site_colors = c("#E6A0C4","#C6CDF7","#D8A499","#7394D4") 

```

# PCoAs

## All samples
### Sex site
```{r}
## BRAY
### Calculate the PCoA on this distance matrix
rt.pcoa = ordinate(physeq_debKno0_taxfix, method="PCoA", distance=bray_all)
plot_scree(rt.pcoa, "Screen plot")
## PCoA ordination
## Define shapes
pcoa<-0
pcoa <- plot_ordination(physeq_debKno0_taxfix, rt.pcoa, "samples", color="site_sex") + stat_ellipse(level = 0.90,type = "t",aes(group=site_sex)) + ggtheme + geom_point(size = 2) + scale_color_manual(values = sex_site_colors)
pcoa
pdf("output/PCoA_bray_all_sitesex.pdf",width=9,height=5)
pcoa
dev.off()

# UUNI
##Calculate the PCoA on this distance matrix
rt.pcoa = ordinate(physeq_debKno0_taxfix, method="PCoA", distance=uuni_all)
plot_scree(rt.pcoa, "Screen plot")
## PCoA ordination
## Define shapes
pcoa<-0
pcoa <- plot_ordination(physeq_debKno0_taxfix, rt.pcoa, "samples", color="site_sex") + stat_ellipse(level = 0.90,type = "t",aes(group=site_sex)) + ggtheme + geom_point(size = 2) + scale_color_manual(values = sex_site_colors)
pcoa
pdf("output/PCoA_uuni_all_sitesex.pdf",width=9,height=5)
pcoa
dev.off()

# WUNI
##Calculate the PCoA on this distance matrix
rt.pcoa = ordinate(physeq_debKno0_taxfix, method="PCoA", distance=wuni_all)
plot_scree(rt.pcoa, "Screen plot")
## PCoA ordination
## Define shapes
pcoa<-0
pcoa <- plot_ordination(physeq_debKno0_taxfix, rt.pcoa, "samples", color="site_sex") + stat_ellipse(level = 0.90,type = "t",aes(group=site_sex)) + ggtheme + geom_point(size = 2) + scale_color_manual(values = sex_site_colors)
pcoa
pdf("output/PCoA_wuni_all_sitesex.pdf",width=9,height=5)
pcoa
dev.off()
```

# BAR PLOTS

## All samples
### Convert to correct file format
```{r}
meco_all_dataset <- phyloseq2meco(physeq_debKno0_taxfix)

```

### Plot bar
```{r}
# create trans_abund object
# select top 15 abundant Phyla.
t1 <- trans_abund$new(dataset = meco_all_dataset, taxrank = "Family", ntaxa = 15)
# t1 object now include the transformed abundance data t1$abund_data and other elements for the following plotting

# return a ggplot2 object
t1$plot_bar(others_color = "grey70", facet = "sex", xtext_keep = FALSE, legend_text_italic = FALSE)

# Two or more facets are supported with the facet parameter from v0.14.0 by providing a vector with multiple elements.
# require package ggh4x, first run install.packages("ggh4x") if not installed
p <- t1$plot_bar(others_color = "grey70", facet = c("sex", "collection_method"), xtext_keep = FALSE, legend_text_italic = FALSE, barwidth = 1)
pdf("output/bar_microeco_all_sexsite_fam_16.pdf",width=9,height=5)
p
dev.off()


###################
# No rick
meco_norick_dataset <- phyloseq2meco(no_rick_prune)

# create trans_abund object
# select top 15 abundant Phyla.
t1 <- trans_abund$new(dataset = meco_norick_dataset, taxrank = "Family", ntaxa = 15)
# t1 object now include the transformed abundance data t1$abund_data and other elements for the following plotting

# return a ggplot2 object
t1$plot_bar(others_color = "grey70", facet = "sex", xtext_keep = FALSE, legend_text_italic = FALSE)

# Two or more facets are supported with the facet parameter from v0.14.0 by providing a vector with multiple elements.
# require package ggh4x, first run install.packages("ggh4x") if not installed
p <- t1$plot_bar(others_color = "grey70", facet = c("sex", "collection_method"), xtext_keep = FALSE, legend_text_italic = FALSE, barwidth = 1)
pdf("output/bar_microeco_norick_sexsite_fam_16.pdf",width=9,height=5)
p
dev.off()
```


# SAMPLE DENDROGRAM

# Generate dendrogram for importing into Evolview
```{r}
# uuni_all
hClusteringuuni <- hclust(uuni_all, method = 'complete')
plot(hClusteringuuni, hang = -1)

#bray_all
hClusteringbray <- hclust(bray_all, method = 'complete')
plot(hClusteringbray, hang = -1)

library(ape)
class(hClusteringuuni) # must be hclust class
my_dend <- as.phylo(hClusteringuuni) 
write.tree(phy=my_dend, file="all_uuni_dend.newick")

library(ape)
class(hClusteringbray) # must be hclust class
my_dend_bray <- as.phylo(hClusteringbray) 
write.tree(phy=my_dend_bray, file="all_bray_dend.newick")


```








