---
title: "Tick16S_Analysis_Pipeline_MASamples_20230810"
author: "Sofia Roitman"
date: "2023-08-10"
output: html_document
---

# Set working directory
```{r}
setwd("/path/to/wd")
```


# Load dependencies
```{r message=FALSE, error=FALSE}
# libraries and functions
library(tidyverse)
library(RVAideMemoire)
library(vegan)
library(BiocManager)
library(phyloseq) 
library(rhdf5)
library(biomformat)
library(microbiome)
library(dplyr) 
library(gcookbook)
library(tidyverse)
library(ape)
library(Biostrings) 
library(Hmisc) 
library(yaml) 
library(tidyr) 
library(stats) 
library(utils) 
library(qiime2R)
library(genefilter)
library(magrittr)
library(DESeq2)
library(plyr)
library(data.table)
library(phytools)
library(stringi)
library(ggpubr)
library(vegan)
library(microbiomeSeq)
library(microbiomeutilities)
library(ampvis2)
library(pairwiseAdonis)
library(ecole)
library(paletteer)
library(mirlyn)
library(ggtree)
library(ggplot2)
library(scales)
library(microbiomeViz)
library(miaViz)
library(scater)
library(mia)
library(mina)
library(NetCoMi)
library(Matrix)
library(igraph)

```

# Import starting phyloseq object
```{r}
readRDS(file = "physeq_debkno0_taxfix.rds")
```


# SUBSETTING

# MA
```{r}
ma_subset = subset_samples(physeq_debKno0_taxfix, collection_method == "flag")
ma <- prune_taxa(taxa_sums(ma_subset) >= 1, ma_subset)
```

## By sex
```{r}

maf = subset_samples(ma, sex == "F")
mam = subset_samples(ma, sex == "M")

maf_prune <- prune_taxa(taxa_sums(maf) > 1, maf)
maf_prune_rel  = transform_sample_counts(maf_prune, function(x) x / sum(x) )


```

### Female only, individual diseases, no coinfections
```{r}
maf_lyme_nocoinf = subset_samples(maf_prune, bm_babesiosis_noPN == "NN" & aph_anaplasmosis_noPN == "NN")
maf_lyme_nocoinf_prune <- prune_taxa(taxa_sums(maf_lyme_nocoinf) > 1, maf_lyme_nocoinf)
maf_lyme_nocoinf_prune_rel  = transform_sample_counts(maf_lyme_nocoinf_prune, function(x) x / sum(x) )

maf_bab_nocoinf = subset_samples(maf_prune, bb_lyme_noPN == "NN" & aph_anaplasmosis_noPN == "NN")
maf_bab_nocoinf_prune <- prune_taxa(taxa_sums(maf_bab_nocoinf) > 1, maf_bab_nocoinf)

maf_ana_nocoinf = subset_samples(maf_prune, bb_lyme_noPN == "NN" & bm_babesiosis_noPN == "NN")
maf_ana_nocoinf_prune <- prune_taxa(taxa_sums(maf_ana_nocoinf) > 1, maf_ana_nocoinf)
```

## NN across all diseases
```{r}

mann = subset_samples(ma, bb_lyme_noPN == "NN" & bm_babesiosis_noPN == "NN" & aph_anaplasmosis_noPN == "NN")

```

## No Rickettsia
```{r}

manorick = subset_taxa(ma, Genus!="Rickettsia")

```


## LYME MA
```{r}
# Lyme PP no PN
malymeppnoPN = subset_samples(ma, bb_lyme_noPN == "PP")
## Lyme PP ONLY
malymepponlynoPN = subset_samples(malymeppnoPN, bm_babesiosis_noPN == "NN" & aph_anaplasmosis_noPN == "NN")
## Lyme PP, Babesiosis PP
malymebabppnoPN = subset_samples(malymeppnoPN, bm_babesiosis_noPN == "PP" & aph_anaplasmosis_noPN == "NN")
## Lyme PP, Anaplasmosis PP
malymeanappnoPN = subset_samples(malymeppnoPN, bm_babesiosis_noPN == "NN" & aph_anaplasmosis_noPN == "PP")
## Lyme PP, Babesiosis PP, Anaplasmosis PP
malymebabanappnoPN = subset_samples(malymeppnoPN, bm_babesiosis_noPN == "PP" & aph_anaplasmosis_noPN == "PP")


# Lyme NN
malymenn = subset_samples(ma, bb_lyme_noPN == "NN")
## Lyme PP and NN ONLY
malymeppnnonly = subset_samples(ma, bm_babesiosis_noPN == "NN" & aph_anaplasmosis_noPN == "NN")

```

## BABESIOSIS MA
```{r}
# Babesiosis PP no PN
mababppnoPN = subset_samples(ma, bm_babesiosis_noPN == "PP")
## Babesiosis PP ONLY
mababpponlynoPN = subset_samples(mababppnoPN, bb_lyme_noPN == "NN" & aph_anaplasmosis_noPN == "NN")
## Babesiosis PP, Anaplasmosis PP
#babanapp = subset_samples(mababppnoPN, bb_lyme_noPN == "NN" & aph_anaplasmosis_noPN == "PP")
#### No co-occurrence of babesiosis and anaplasmosis!

# Babesiosis NN
mababnn = subset_samples(ma, bm_babesiosis_noPN == "NN")
## Babesiosis PP and NN ONLY
mababppnnonly = subset_samples(ma, bb_lyme_noPN == "NN" & aph_anaplasmosis_noPN == "NN")

```

## ANAPLASMOSIS
```{r}
# Anaplasmosis PP no PN
maanappnoPN = subset_samples(ma, aph_anaplasmosis_noPN == "PP")
## Anaplasmosis PP ONLY
maanapponlynoPN = subset_samples(maanappnoPN, bb_lyme_noPN == "NN" & bm_babesiosis_noPN == "NN")


# Anaplasmosis NN
maanann = subset_samples(ma, aph_anaplasmosis_noPN == "NN")
## Anaplasmosis PP and NN ONLY
maanappnnonly = subset_samples(ma, bb_lyme_noPN == "NN" & bm_babesiosis_noPN == "NN")

```

# DISTANCE MATRICES
## Bray Curtis and Unweighted Unifrac

### Basic unsubsetted dataset
```{r}
# All
bray_ma <- phyloseq::distance(ma, "bray")
uuni_ma <- phyloseq::distance(ma, "uUniFrac")
wuni_ma <- phyloseq::distance(ma, "wUniFrac")

```

## NN across all diseases
```{r}

bray_mann <- phyloseq::distance(mann, "bray")
uuni_mann <- phyloseq::distance(mann, "uUniFrac")
wuni_mann <- phyloseq::distance(mann, "wUniFrac")

```

## No Rickettsia
```{r}

bray_manorick <- phyloseq::distance(manorick, "bray")
uuni_manorick <- phyloseq::distance(manorick, "uUniFrac")
wuni_manorick <- phyloseq::distance(manorick, "wUniFrac")

```


### Sex
```{r}
bray_maf_prune <- phyloseq::distance(maf_prune, "bray")
uuni_maf_prune <- phyloseq::distance(maf_prune, "uUniFrac")
wuni_maf_prune <- phyloseq::distance(maf_prune, "wUniFrac")

bray_maf_lyme_nocoinf_prune <- phyloseq::distance(maf_lyme_nocoinf_prune, "bray")
uuni_maf_lyme_nocoinf_prune <- phyloseq::distance(maf_lyme_nocoinf_prune, "uUniFrac")
wuni_maf_lyme_nocoinf_prune <- phyloseq::distance(maf_lyme_nocoinf_prune, "wUniFrac")

# Field
bray_mam <- phyloseq::distance(mam, "bray")
uuni_mam <- phyloseq::distance(mam, "uUniFrac")
wuni_mam <- phyloseq::distance(mam, "wUniFrac")
```

#### Female, individual diseases, no coinfections
```{r}
bray_maf_lyme_nocoinf_prune <- phyloseq::distance(maf_lyme_nocoinf_prune, "bray")
uuni_maf_lyme_nocoinf_prune <- phyloseq::distance(maf_lyme_nocoinf_prune, "uUniFrac")
wuni_maf_lyme_nocoinf_prune <- phyloseq::distance(maf_lyme_nocoinf_prune, "wUniFrac")

bray_maf_bab_nocoinf_prune <- phyloseq::distance(maf_bab_nocoinf_prune, "bray")
uuni_maf_bab_nocoinf_prune <- phyloseq::distance(maf_bab_nocoinf_prune, "uUniFrac")
wuni_maf_bab_nocoinf_prune <- phyloseq::distance(maf_bab_nocoinf_prune, "wUniFrac")

bray_maf_ana_nocoinf_prune <- phyloseq::distance(maf_ana_nocoinf_prune, "bray")
uuni_maf_ana_nocoinf_prune <- phyloseq::distance(maf_ana_nocoinf_prune, "uUniFrac")
wuni_maf_ana_nocoinf_prune <- phyloseq::distance(maf_ana_nocoinf_prune, "wUniFrac")
```


### Lyme
```{r}
# Lyme PP noPN
bray_malymepp_noPN <- phyloseq::distance(malymeppnoPN, "bray")
uuni_malymepp_noPN <- phyloseq::distance(malymeppnoPN, "uUniFrac")
wuni_malymepp_noPN <- phyloseq::distance(malymeppnoPN, "wUniFrac")
## Lyme PP ONLY
bray_malymepponly_noPN <- phyloseq::distance(malymepponlynoPN, "bray")
uuni_malymepponly_noPN <- phyloseq::distance(malymepponlynoPN, "uUniFrac")
wuni_malymepponly_noPN <- phyloseq::distance(malymepponlynoPN, "wUniFrac")
## Lyme PP, Babesiosis PP
bray_malymebabpp_noPN <- phyloseq::distance(malymebabppnoPN, "bray")
uuni_malymebabpp_noPN <- phyloseq::distance(malymebabppnoPN, "uUniFrac")
wuni_malymebabpp_noPN <- phyloseq::distance(malymebabppnoPN, "wUniFrac")
## Lyme PP, Anaplasmosis PP
bray_malymeanapp_noPN <- phyloseq::distance(malymeanappnoPN, "bray")
uuni_malymeanapp_noPN <- phyloseq::distance(malymeanappnoPN, "uUniFrac")
wuni_malymeanapp_noPN <- phyloseq::distance(malymeanappnoPN, "wUniFrac")
## Lyme PP, Babesiosis PP, Anaplasmosis PP
bray_malymebabanapp_noPN <- phyloseq::distance(malymebabanappnoPN, "bray")
uuni_malymebabanapp_noPN <- phyloseq::distance(malymebabanappnoPN, "uUniFrac")
wuni_malymebabanapp_noPN <- phyloseq::distance(malymebabanappnoPN, "wUniFrac")


# Lyme NN
bray_malymenn <- phyloseq::distance(malymenn, "bray")
uuni_malymenn <- phyloseq::distance(malymenn, "uUniFrac")
wuni_malymenn <- phyloseq::distance(malymenn, "wUniFrac")
## Lyme PP and NN ONLY
bray_malymeppnnonly <- phyloseq::distance(malymeppnnonly, "bray")
uuni_malymeppnnonly <- phyloseq::distance(malymeppnnonly, "uUniFrac")
wuni_malymeppnnonly <- phyloseq::distance(malymeppnnonly, "wUniFrac")
```

### Babesiosis
```{r}
# Babesiosis PP noPN
bray_mababppnoPN <- phyloseq::distance(mababppnoPN, "bray")
uuni_mababppnoPN <- phyloseq::distance(mababppnoPN, "uUniFrac")
wuni_mababppnoPN <- phyloseq::distance(mababppnoPN, "wUniFrac")
## Babesiosis PP ONLY
bray_mababpponlynoPN <- phyloseq::distance(mababpponlynoPN, "bray")
uuni_mababpponlynoPN <- phyloseq::distance(mababpponlynoPN, "uUniFrac")
wuni_mababpponlynoPN <- phyloseq::distance(mababpponlynoPN, "wUniFrac")

# Babesiosis NN
bray_mababnn <- phyloseq::distance(mababnn, "bray")
uuni_mababnn <- phyloseq::distance(mababnn, "uUniFrac")
wuni_mababnn <- phyloseq::distance(mababnn, "wUniFrac")
## Babesiosis PP and NN ONLY
bray_mababppnnonly <- phyloseq::distance(mababppnnonly, "bray")
uuni_mababppnnonly <- phyloseq::distance(mababppnnonly, "uUniFrac")
wuni_mababppnnonly <- phyloseq::distance(mababppnnonly, "wUniFrac")
```

### Anaplasmosis
```{r}
# Anaplasmosis PP noPN
bray_maanappnoPN <- phyloseq::distance(maanappnoPN, "bray")
uuni_maanappnoPN <- phyloseq::distance(maanappnoPN, "uUniFrac")
## Anaplasmosis PP ONLY
bray_maanapponlynoPN <- phyloseq::distance(maanapponlynoPN, "bray")
uuni_maanapponlynoPN <- phyloseq::distance(maanapponlynoPN, "uUniFrac")

# Anaplasmosis NN
bray_maanann <- phyloseq::distance(maanann, "bray")
uuni_maanann <- phyloseq::distance(maanann, "uUniFrac")
## Anaplasmosis PP and NN ONLY
bray_maanappnnonly <- phyloseq::distance(maanappnnonly, "bray")
uuni_maanappnnonly <- phyloseq::distance(maanappnnonly, "uUniFrac")
```

# Set theme: Use ggplot theme to make everything more legible
```{r}
ggtheme <- theme(axis.title = element_text(colour="black",family = "Helvetica",
                                           size = rel(1.5)), 
                 axis.text = element_text(family = "Helvetica",colour = "black",
                                          size = rel(1)), 
                 axis.line = element_line(size = 0.5,colour = "black"), 
                 axis.ticks = element_blank(),
                 panel.grid.major = element_line(colour="grey",size = rel(0.25)), 
                 panel.grid.minor = element_blank(), 
                 panel.background = element_blank(),
                 plot.title = element_text(colour = "black", face = "bold",
                                           size = rel(2),family = "Helvetica",hjust = 0.5,
                 geom_point(size = 20)))

lyme_colors = c("#f49c98", "#fcd68f", "#629076")
lyme_colors_noPN = c("#f49c98", "#fcd3d2")
ana_colors_noPN = c("#8f6aa3","#d2c1db")
bab_colors_noP0N = c("#629076","#cae8d7")
sex_colors = c("#F78FB2","#3E8AC6")
region_colors = c("#fcd68f",)
```


# SIGNIFICANCE TESTING

## ALL
### PERMANOVA
```{r}
# ALL
df <- data.frame(sample_data(ma))
distance_methods <-c("bray_ma")
set.seed(129)
# Run for loop in distance matrices         
for (i in distance_methods){ 
  form <- as.formula(paste(i, "sex+bb_lyme_noPN+bm_babesiosis_noPN+aph_anaplasmosis_noPN", sep="~"))
  print(form)
 adonis2(form, data=df)->result 
 print(result)
 capture.output(result, file = paste0("output/","ADONIS_bray_ma_noPN",date(),".txt"))
}

## UUNI
### Make a data frame from the otu_data
df <- data.frame(sample_data(ma))
distance_methods <-c("uuni_ma")
set.seed(129)
# Run for loop in distance matrices         
for (i in distance_methods){ 
  form <- as.formula(paste(i, "sex+bb_lyme_noPN+bm_babesiosis_noPN+aph_anaplasmosis_noPN", sep="~"))
  print(form)
 adonis2(form, data=df)->result
 print(result)
 capture.output(result, file = paste0("output/","ADONIS_uuni_ma_noPN",date(),".txt"))
} 


## WUNI
df <- data.frame(sample_data(ma))
distance_methods <-c("wuni_ma")
set.seed(129)
# Run for loop in distance matrices         
for (i in distance_methods){ 
  form <- as.formula(paste(i, "sex+bb_lyme_noPN+bm_babesiosis_noPN+aph_anaplasmosis_noPN", sep="~"))
  print(form)
 adonis2(form, data=df)->result
 print(result)
 capture.output(result, file = paste0("output/","ADONIS_wuni_ma_noPN",date(),".txt"))
}  
```

#### Posthoc
```{r}
df <- data.frame(sample_data(ma))

# BRAY                 
## Pariwise Permutation MANOVAs: lyme
set.seed(129)
permanova_pairwise(bray_all,df$site_sex,permutations = 1000)



set.seed(129)
permanova_pairwise(bray_ma,df$path_long_noPN,permutations = 1000)
### No sig

# UUNI
## Pariwise Permutation MANOVAs: sex_site
set.seed(129)
permanova_pairwise(uuni_all,df$site_sex,permutations = 1000)


## Pariwise Permutation MANOVAs: anaplasma
set.seed(129)
permanova_pairwise(uuni_ma,df$path_long_noPN,permutations = 1000)




```

### PERMDISP
#### BRAY
```{r}
# Sex
## make a data frame from the sample_data
sampledf <- data.frame(sample_data(ma))
## Make a data frame from the otu_data
beta <- betadisper(bray_ma, sampledf$sex)
result <- permutest(beta, pairwise = TRUE)
print(result)
capture.output(result, file = paste0("output/","BETADISP_bray_ma_sex",date(),".txt"))
### 0.001


# Lyme
## make a data frame from the sample_data
sampledf <- data.frame(sample_data(ma))
## Make a data frame from the otu_data
beta <- betadisper(bray_ma, sampledf$bb_lyme_noPN)
result <- permutest(beta, pairwise = TRUE)
print(result)
capture.output(result, file = paste0("output/","BETADISP_bray_ma_bblyme_noPN",date(),".txt"))
### No sig

# Bab
## make a data frame from the sample_data
sampledf <- data.frame(sample_data(ma))
## Make a data frame from the otu_data
beta <- betadisper(bray_ma, sampledf$bm_babesiosis_noPN)
result <- permutest(beta, pairwise = TRUE)
print(result)
capture.output(result, file = paste0("output/","BETADISP_bray_ma_bmbabesiosis_noPN",date(),".txt"))
### 0.003

# Ana
## make a data frame from the sample_data
sampledf <- data.frame(sample_data(ma))
## Make a data frame from the otu_data
beta <- betadisper(bray_ma, sampledf$aph_anaplasmosis_noPN)
result <- permutest(beta, pairwise = TRUE)
print(result)
capture.output(result, file = paste0("output/,"BETADISP_bray_ma_aphanaplasmosis_noPN",date(),".txt"))
### No sig

```

#### UUNI
```{r}
# Sex
## make a data frame from the sample_data
sampledf <- data.frame(sample_data(ma))
## Make a data frame from the otu_data
beta <- betadisper(uuni_ma, sampledf$sex)
result <- permutest(beta, pairwise = TRUE)
print(result)
capture.output(result, file = paste0("output/betadisp_2023_05/",i,"BETADISP_uuni_ma_sex",date(),".txt"))
## No sig

# Lyme
## noPN
### make a data frame from the sample_data
sampledf <- data.frame(sample_data(ma))
## Make a data frame from the otu_data
beta <- betadisper(uuni_ma, sampledf$bb_lyme_noPN)
result <- permutest(beta, pairwise = TRUE)
print(result)
capture.output(result, file = paste0("output/betadisp_2023_05/",i,"BETADISP_uuni_ma_bblyme_noPN",date(),".txt"))
## No sig

# Bab
## noPN
### make a data frame from the sample_data
sampledf <- data.frame(sample_data(ma))
## Make a data frame from the otu_data
beta <- betadisper(uuni_ma, sampledf$bm_babesiosis_noPN)
result <- permutest(beta, pairwise = TRUE)
print(result)
capture.output(result, file = paste0("output/betadisp_2023_05/",i,"BETADISP_uuni_ma_bmbabesiosis_noPN",date(),".txt"))
### Nosig


# Ana
## noPN
### make a data frame from the sample_data
sampledf <- data.frame(sample_data(ma))
## Make a data frame from the otu_data
beta <- betadisper(uuni_ma, sampledf$aph_anaplasmosis_noPN)
result <- permutest(beta, pairwise = TRUE)
print(result)
capture.output(result, file = paste0("output/betadisp_2023_05/",i,"BETADISP_uuni_ma_aphanaplasmosis_noPN",date(),".txt"))
### No sig

```


# LOGFOLD CHANGE USING DESEQ
## FEMALE
### LYME PP vs. NN ONLY, no co-infections
```{r}


# DESeq- requires absolute counts, not relative proportions
deseqdata<-phyloseq_to_deseq2(maf_nocoinf, ~bb_lyme_noPN)
gm_mean = function(x, na.rm=TRUE){exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))}
geoMeans<-apply(counts(deseqdata),1,gm_mean)
deseqdata<-estimateSizeFactors(deseqdata,geoMeans=geoMeans)
deseqdata<-DESeq(deseqdata,fitType="local")
res<-results(deseqdata,contrast=c("bb_lyme_noPN","PP","NN"))
alpha<-0.05
sigtab<-res[which(res$padj<alpha),]
sigtab<-cbind(as(sigtab,"data.frame"),as(tax_table(maf_nocoinf)[rownames(sigtab),],"matrix"))
head(sigtab)
write.table(sigtab,"output/Logfold_SidDif_MAFnocoinfLyme_PPnum_noPN_2023_07_14.txt",sep="\t")


# Plot
## Plot Results
theme_set(theme_bw())
sigtabgen = subset(sigtab, !is.na(Family))
# Order order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Order, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Order = factor(as.character(sigtabgen$Order), levels=names(x))
# Family order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Family, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Family = factor(as.character(sigtabgen$Family), levels=names(x))
log_lyme <- ggplot(sigtabgen, aes(y=Family, x=log2FoldChange, color=Order)) + 
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_point(size=6) + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5) +
  theme(axis.title.x = element_text(size=14))+
  theme(axis.title.y = element_text(size=14))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12)))


log_lyme

pdf("output/Logfold_MAFnocoinfLymePP_numerator_noPN_2023_07_14_test.pdf", width=10,height=7)
log_lyme
dev.off()

```




### BAB PP vs. NN only, prune, no coinfections
```{r}
# DESeq- requires absolute counts, not relative proportions
deseqdata<-phyloseq_to_deseq2(maf_bab_nocoinf_prune, ~bm_babesiosis_noPN)
gm_mean = function(x, na.rm=TRUE){exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))}
geoMeans<-apply(counts(deseqdata),1,gm_mean)
deseqdata<-estimateSizeFactors(deseqdata,geoMeans=geoMeans)
deseqdata<-DESeq(deseqdata,fitType="local")
res<-results(deseqdata,contrast=c("bm_babesiosis_noPN","PP","NN"))
alpha<-0.05
sigtab<-res[which(res$padj<alpha),]
sigtab<-cbind(as(sigtab,"data.frame"),as(tax_table(maf)[rownames(sigtab),],"matrix"))
head(sigtab)
write.table(sigtab,"output/Logfold_SidDif_MAFprunenocoinfBab_PPnum_noPN_2023_07_17.txt",sep="\t")


# Plot
## Plot Results
theme_set(theme_bw())
sigtabgen = subset(sigtab, !is.na(Family))
# Order order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Order, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Order = factor(as.character(sigtabgen$Order), levels=names(x))
# Family order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Family, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Family = factor(as.character(sigtabgen$Family), levels=names(x))
log_bab <- ggplot(sigtabgen, aes(y=Family, x=log2FoldChange, color=Order)) + 
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_point(size=6) + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5) +
  theme(axis.title.x = element_text(size=14))+
  theme(axis.title.y = element_text(size=14))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12)))


log_bab

pdf("output/Logfold_MAFprunenocoinfBabPP_numerator_noPN_2023_07_17.pdf")
log_bab
dev.off()

```

## Ana PP vs. NN only, prune, no coinfections
```{r}
# DESeq- requires absolute counts, not relative proportions
deseqdata<-phyloseq_to_deseq2(maf_ana_nocoinf_prune, ~aph_anaplasmosis_noPN)
gm_mean = function(x, na.rm=TRUE){exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))}
geoMeans<-apply(counts(deseqdata),1,gm_mean)
deseqdata<-estimateSizeFactors(deseqdata,geoMeans=geoMeans)
deseqdata<-DESeq(deseqdata,fitType="local")
res<-results(deseqdata,contrast=c("aph_anaplasmosis_noPN","PP","NN"))
alpha<-0.05
sigtab<-res[which(res$padj<alpha),]
sigtab<-cbind(as(sigtab,"data.frame"),as(tax_table(ma)[rownames(sigtab),],"matrix"))
head(sigtab)
write.table(sigtab,"output/Logfold_SidDif_MAFprunenocoinfAna_PPnum_noPN_2023_07_17.txt",sep="\t")


# Plot
## Plot Results
theme_set(theme_bw())
sigtabgen = subset(sigtab, !is.na(Family))
# Order order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Order, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Order = factor(as.character(sigtabgen$Order), levels=names(x))
# Family order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Family, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Family = factor(as.character(sigtabgen$Family), levels=names(x))
log_ana <- ggplot(sigtabgen, aes(y=Family, x=log2FoldChange, color=Order)) + 
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_point(size=6) + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5) +
  theme(axis.title.x = element_text(size=14))+
  theme(axis.title.y = element_text(size=14))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12)))


log_ana

pdf("output/Logfold_MAFprunenocoinfAnaPP_numerator_noPN_2023_07_17.pdf")
log_ana
dev.off()

```


# Creating co-occurrence networks
## Visualizing in Cytoscape
### MAF Lyme-negative
```{r}
#maf_prune_genglom <- tax_glom(maf_prune, taxrank = "Genus")
maf_genglom_nn <- subset_samples(maf_prune_genglom, path_long_noPN == "NN")
maf_prune_genglom_nn <- prune_taxa(taxa_sums(maf_genglom_nn) > 1, maf_genglom_nn)
write.csv(otu_table(maf_prune_genglom_nn), file = "maf_prune_genglom_nn_asvtab_20230808.csv")
write.csv(tax_table(maf_prune_genglom_nn), file = "maf_prune_genglom_nn_taxtab_20230808.csv")
otu.mafnn.tab <- read.csv(file = "maf_prune_genglom_nn_asvtab_20230808.csv", header=T, row.names = 1)
tax.mafnn.tab <- read.csv(file = "maf_prune_genglom_nn_taxtab_20230808.csv", header = T, row.names = 1)

otu.mafnn.tab.t <- transpose(otu.mafnn.tab)
rownames(otu.mafnn.tab.t) <- colnames(otu.mafnn.tab)
colnames(otu.mafnn.tab.t) <- tax.mafnn.tab$Genus
tax.mafnn.tab.test <- tax.mafnn.tab
rownames(tax.mafnn.tab.test) <- tax.mafnn.tab$Genus

# Calculating the “Spearman” correlation coefficient between OTUs using the function rcorr(). A list will be created from, with three elements, r for correlations, n for the number of observations and P for the p-values. [If you are curious type ?rcorr() to explore the coefficient options].
otu.mafnn.cor <- rcorr(as.matrix(otu.mafnn.tab.t), type = "pearson")
otu.mafnn.cor2 <- otu.mafnn.cor
#diag(otu.cor2$r) <- 0
# To obtain the p-value information from the list, use otu.cor$P, using the function forceSymmetric() to assign self-correlation as NA.
#otu.pval <- forceSymmetric(otu.cor$P)

# Select only the taxa for the filtered OTUs by using rownames of otu.pval
#sel.tax <- tax.tab.test[rownames(otu.pval),,drop = FALSE]
# Sanity check
#all.equal(rownames(sel.tax), rownames(otu.pval))

# Filter the association based on p-values and level of correlations
#p.yes <- otu.pval<0.05

# Select the r values for the filter probality of < 0.5.
#r.val = otu.cor$r # select all the correlation values 
#p.yes.r <- r.val*p.yes # only select correlation values based on p-value criterion 

# Select OTUs by level of correlation
r_otucor.mafnn <- otu.mafnn.cor2$r
r75.mafnn <- abs(r_otucor.mafnn)>0.75 # output is logical vector
p.yes.rr.mafnn <- r_otucor.mafnn*r75.mafnn # use logical vector for subscripting.

# Select only the taxa for the filtered OTUs by using rownames of otu.pval
sel.mafnn.tax <- tax.mafnn.tab.test[rownames(p.yes.rr.mafnn),,drop = FALSE]
# Sanity check
all.equal(rownames(sel.mafnn.tax), rownames(p.yes.rr.mafnn))

#Create an adjacency matrix
adjm.mafnn <- as.matrix(p.yes.rr.mafnn)

# Add taxonomic information from the metadata associated with adjacency matrix
sel.mafnn.tax2 <- sel.mafnn.tax
sel.mafnn.tax2$ASV <- rownames(sel.mafnn.tax)
colnames(adjm.mafnn) <- as.vector(sel.mafnn.tax2$Genus)
rownames(adjm.mafnn) <- as.vector(sel.mafnn.tax2$Genus)

# iGraph can create graph objects from different sources like data.frame, edge.list, or from adjacency.matrix.
# The function graph.adjacency creates a graph object from adjacency matrix. Weight here represents the level of correlation.
net.grph=graph.adjacency(adjm.mafnn,mode="undirected",weighted=TRUE,diag=FALSE)

#Obtaining edge weight based on the Spearman correlation
edgew<-E(net.grph)$weight

# Creating a vector to remove the isolated nodes (nodes with no interactions)
bad.vs<-V(net.grph)[degree(net.grph) == 0] 

# Removing the isolated nodes from the graph object using the function delete.vertices()
net.grph <-delete.vertices(net.grph, bad.vs)

# Simplfy the adjacency object
g <- simplify(net.grph, remove.multiple=TRUE, remove.loops=TRUE)

# Colour negative correlation edges as blue
E(g)[which(E(g)$weight<0)]$color <- "darkred"

# Colour positive correlation edges as red
E(g)[which(E(g)$weight>0)]$color <- "darkgreen"

# Convert edge weights to absolute values
E(g)$weight <- abs(E(g)$weight)

# Remove edges below absolute Pearson correlation 0.8
g <- delete_edges(g, E(g)[which(E(g)$weight<0.8)])

# Remove any vertices remaining that have no edges
g <- delete.vertices(g, degree(g)==0)

g2 <- g
E(g2)$r <- otu.mafnn.cor2$r 
plotCytoscapeGML(g, file = "maf_specglom_igraph_to_cytoscape_g.gml")

plot.igraph(g, vertex.size=20, edge.width=abs(E(g)$weight)*8, 
        edge.color=ifelse(otu.mafnn.cor2$r > 0, "blue","red"))

plot(net.grph,
    vertex.size=8,
    vertex.frame.color="black",
    edge.curved=F,
    edge.width=1.5,
    layout=layout.fruchterman.reingold,
    edge.color=ifelse(edgew < 0,"red","blue"),
    vertex.label=NA,
    vertex.label.color="black",
    vertex.label.family="Times New Roman",
    vertex.label.font=2)

net.grph$edgew
edgew

library(RCy3)
createNetworkFromIgraph(net.grph, title = "maf_igraph_pearson", collection = "My first iGraph")

library(NetPathMiner)
plotCytoscapeGML(net.grph, file = "mafnn_genglom_igraph_to_cytoscape_20230808.gml")
```

## MAF Lyme-positive
```{r}
#maf_prune_genglom <- tax_glom(maf_prune, taxrank = "Genus")
maf_genglom_lyme <- subset_samples(maf_prune_genglom, path_long_noPN == "lyme")
maf_prune_genglom_lyme <- prune_taxa(taxa_sums(maf_genglom_lyme) > 1, maf_genglom_lyme)
maf_prune_genglom_lyme_rel  = transform_sample_counts(maf_prune_genglom_lyme, function(x) x / sum(x) )
write.csv(otu_table(maf_prune_genglom_lyme_rel), file = "maf_prune_rel_genglom_lyme_asvtab_20230808.csv")
write.csv(tax_table(maf_prune_genglom_lyme_rel), file = "maf_prune_rel_genglom_lyme_taxtab_20230808.csv")
otu.maflyme.tab <- read.csv(file = "maf_prune_rel_genglom_lyme_asvtab_20230808.csv", header=T, row.names = 1)
tax.maflyme.tab <- read.csv(file = "maf_prune_rel_genglom_lyme_taxtab_20230808.csv", header = T, row.names = 1)

otu.maflyme.tab.gen <- otu.maflyme.tab
otu.maflyme.tab.gen$genus <- tax.maflyme.tab$Genus

sp_ratio.maflyme <- otu.maflyme.tab.gen %>% 
  as_tibble() %>%
  rename_all(tolower) %>%
  column_to_rownames(var = "genus")

min.prevalence=10
incidence=sp_ratio.maflyme
incidence[incidence>0]=1
sp_ratio_filtered.maflyme <- sp_ratio.maflyme[which(rowSums(incidence)>=min.prevalence),] ### end of prevalence filtering

sp_correl.maflyme <- sp_ratio_filtered.maflyme %>% 
  t() %>% 
  cor(method = "pearson")  ### correlation calculations

sp_correl.maflyme[abs(sp_correl.maflyme)<0.7]=0  ### define threshold for correlations

net_work.maflyme <- graph_from_adjacency_matrix(sp_correl.maflyme,mode="undirected",weighted=TRUE, diag=FALSE)  ## create network file
net_work.maflyme <- delete.vertices(net_work.maflyme,degree(net_work.maflyme)==0) #remove nodes without edges

plot(net_work.maflyme, vertex.label = NA, edge.width = 5, vertex.size = 10) ## first plot 

plotCytoscapeGML(net_work.maflyme, file = "maflyme_genglom_igraph_to_cytoscape_attributes0.gml")

# Now we will enhance the network appearance by adding attributes to the nodes and edges. It will also make the plot much more readable and will help us during interpretations. Each major steps of the code will be explained.

taxa_tbl.maflyme <- tax.maflyme.tab %>% 
  as_tibble() %>% 
  rename_all(tolower) #tidy taxonomic infos

net_work_used.maflyme <- V(net_work.maflyme)$name %>% 
   as_tibble() %>% 
  mutate(genus = value) %>% 
  select(genus) #extract genera represented in the network

v_attr.maflyme <- sp_ratio.maflyme %>% ### we create a table of attributes for nodes (vertex)
  rownames_to_column( var = "genus") %>% 
  as_tibble() %>% 
  pivot_longer(-genus, names_to = "sample_id", values_to = "ratio" ) %>%
  group_by(genus) %>%
  dplyr::summarise(rel_abun = sum(ratio)) %>% 
  inner_join(net_work_used.maflyme, by = "genus") %>%
  inner_join(taxa_tbl.maflyme, by = "genus") %>%
  mutate(rel_abun = abs(exp (rel_abun)))

network_table.maflyme <- igraph::as_data_frame(net_work.maflyme, 'both') ##we convert the network in data frames
     
network_table.maflyme$vertices <- network_table.maflyme$vertices %>%
    as_tibble() %>% 
    inner_join(v_attr.maflyme, by = c("name"="genus"))  # we add our attribute in the data frames 
  
  net_work1.maflyme <- graph_from_data_frame(network_table.maflyme$edges,
                                     directed = F,
                                     vertices = network_table.maflyme$vertices) # we convert back data frames to a network
plotCytoscapeGML(net_work1.maflyme, file = "maflyme_genglom_igraph_to_cytoscape_attributes0.5.gml")


mom_data.maflyme <- V(net_work1.maflyme)$family %>% 
  as_tibble_col(column_name = "genus") #formating the class variable as factor (is needed for coloring edges)
mom_data.maflyme$genus <- as_factor(mom_data.maflyme$genus)

color_easy <-  c("pink", "brown", "orange", "dodgerblue", "lightgray")[mom_data.maflyme$species] #creating a color palette to represent each class levels

V(net_work1.maflyme)$color <-  color_easy ## we have now the color attributes based on class

E(net_work1.maflyme)$sign <- E(net_work1.maflyme)$weight ##we create another attribute from the weight

E(net_work1.maflyme)$weight <- abs(E(net_work1.maflyme)$weight) ## we then use absolute value of weight because of the specific layout we will use

library(NetPathMiner)
plotCytoscapeGML(net_work1.maflyme, file = "maflyme_genglom_igraph_to_cytoscape_attributes1.gml")
```




